---
title: "EDA and Descriptives in R for AARUG Presentation"
author: "Barry DeCicco"
date: "For Ann Arbor R Users' Group, Mar 9, 2023"
output:  
  html_document:
    toc:  true
    toc_float: true
    code_folding: show



  
---
```{r}

  # ioslides_presentation:
  #   template: quarterly-report.html

```

```{r setup,  collapse=TRUE}  

# load libraries as needed:
# 

library(pastecs)
# https://stats.idre.ucla.edu/r/faq/how-can-i-get-a-table-of-basic-descriptive-statistics-for-my-variables/

# For the gapminder data sets:
# install.packages("gapminder")

library(htmlTable)
library(psych)
library(tidyverse)
library(inspectdf)
library(janitor)
library(skimr)
library(DataExplorer)
library(inspectdf)   
# https://www.r-bloggers.com/part-2-simple-eda-in-r-with-inspectdf/
library(Hmisc)
library(summarytools)
library(readr)

# library(devtools)
# devtools::install_github("alastairrushworth/inspectdf")


 
library(kableExtra)
library(magrittr)

library(gapminder)

library(naniar)

```

## Import the Gapminder data set:


```{r import data, include=TRUE}


View(gapminder)

```

# Using Base R Commands to Inspect the Data Frame

This increases your knowledge of the data set, and catches import errors.

```{r eda, collapse=TRUE}
dim(gapminder)
class(gapminder)
lapply(gapminder, class)
sapply(gapminder, function(x) sum(is.na(x)))   
sapply(gapminder, function(x) sum(is.na(x)/length(x)))  
# https://sebastiansauer.github.io/sum-isna/
str(gapminder)
summary(gapminder)
head(gapminder)
tail(gapminder)

```


## Visualizing Missing Values:

```{r vissualize-missing-values}

gapminder %>%
  vis_miss()


```


## The 'lapply' command can carry out a function across a list. A dataframe is a list of columns:

```{r}

lapply(gapminder, class)

```

### Sometimes you want to clean this up - use the <strong>kable</strong> command:


```{r lapply-class-kable}

gapminder %>%
  lapply(., class) %>%
  as.data.frame() %>%
  pivot_longer(cols=everything(),
               names_to='Variable',
               values_to='Type') %>%
  dplyr::arrange(Type, Variable ) %>%
  kable(caption = "<strong><center>Variable Classes</center></strong>", full_width = F) %>%
  # pack_rows(index = c("String Variable" = 4, "Date" = 1,
  #                     "Numeric" = 4)) %>%
  column_spec(1, width = "5cm", bold = T) %>%
  column_spec(2, width = "5cm") 


```



## Summary table using <strong>pastecs</strong> package:

```{r pastecs, collapse=TRUE}



# Set formatting options: 

options(scipen=100)
options(digits=2)

gapminder %>% 
  select_if(is.numeric) %>% 
  stat.desc()





```



## Summary table with fewer statistics:

```{r pastecs-fewer, collapse=TRUE}



# Set formatting options: 

options(scipen=100)
options(digits=2)

# Run a more limited set of statistics:

gapminder %>% 
  select_if(is.numeric) %>% 
  stat.desc(., basic=F)




```


## Getting a better table with the <strong>kable</strong> package:

First, convert the row names to a column.


```{r pastecs-then-kable, collapse=TRUE}



# Set formatting options: 

options(scipen=100)
options(digits=2)

gapminder %>% 
  select_if(is.numeric) %>% 
  stat.desc() %>% 
  tibble::rownames_to_column(var="Statistic") %>% 
  kable(caption = "<strong><center>Descriptive Statistics</center></strong>", full_width = F,
        escape = FALSE,
   col.names= c("Statistic <br/> Name",
                "Year",
                "Life Expectancy",
                "Population",
                "GDP per<br/>Capita" )
) %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>% 
  column_spec(1, width = "5cm", bold = T) %>%
  column_spec(2:5, width = "3cm") 



```


# The <strong>hmisc::describe</strong> packages

There are two describe commands, one in the Hmisc pacakge, and one in the psych
package.  We will use the <strong>Hmisc</strong> version


## Hmisc Describe


Running the command with no formatting.  It's ugly.


```{r Hmisc-describe, collapse=TRUE}


gapminder %>%
  Hmisc::describe()


```


Cleaning up the output with the <strong>html()</strong> function:


```{r Hmisc-describe-then-html}


s <- gapminder %>%
  Hmisc::describe() 

Hmisc::html(s, title = 'Using HTML Function')



```

## Hmisc::summaryM

This gives a nice table with statistical tests

<strong>Switching to the mtcars data set for this:</strong>

```{r Hmisc-summaryM-by-group}



s <-
  Hmisc::summaryM(
 mpg +
 disp +
 hp + 
 drat + 
 wt +
 qsec +
 vs +
 am + 
 gear +
 carb
           ~ cyl,
    data = mtcars,
    overall = TRUE,
    test = TRUE
  )


Hmisc::html(
  s,
  caption = 'Descriptive Statistics Cylinder, mtcars data',
  exclude1 = TRUE,
  npct = 'both',
  digits = 3,
  prmsd = TRUE,
  brmsd = TRUE,
  msdsize = mu$smaller2
)



```


# Packages for exploring a df:

There are several packages for quick and easy EDA:

*   inspectdf (https://www.littlemissdata.com/blog/inspectdf)<p>
*    skimr (https://cmdlinetips.com/2018/03/skimr-a-r-package-to-skim-summary-data-effortlessly/, 
https://www.datanovia.com/en/blog/display-a-beautiful-summary-statistics-in-r-using-skimr-package/)<p>
*   DataExplorer (https://www.programmingwithr.com/how-to-automate-eda-with-dataexplorer-in-r/)<p>

## Skimr

Plain - you can run it through the kable() function, as above.
Running it through 'as.dataframe()' lets you use all data manipulation tools.

```{r skimr-plain}

skim(gapminder)


```

Choose your own functions, add grouping:
  
```{r skimr-custom}

gapminder %>% 
  filter(year==2007) %>%
  select(continent,lifeExp,gdpPercap)%>%
  group_by(continent) %>%
  skim()


```


## Inspectdf:


Plain:


```{r inspectdf-plain}

inspect_types(gapminder)  %>% show_plot()

```


Missing values:


```{r inspectdf-NA}

inspect_na(gapminder)  %>% show_plot()

```

Numeric variables:

```{r inspectdf-num}

inspect_num(gapminder)  %>% show_plot()

```

## DataExplorer:

Introduce:

```{r DataExplorer-introduce}

introduce(gapminder)
```
Plot missingness:

```{r DataExplorer-plot_missing}

plot_missing(gapminder)

```

# References:


This document is at:  https://github.com/AnnArborRUserGroup/AARUG_EDA_Descriptives_2023_03_09
If it moves, search the repository for the file name.


# Packages for exploring a df:


<strong> There are several packages for quick and easy EDA:</strong><p>


*   inspectdf (https://www.littlemissdata.com/blog/inspectdf)<p>
*    skimr (https://cmdlinetips.com/2018/03/skimr-a-r-package-to-skim-summary-data-effortlessly/, 
https://www.datanovia.com/en/blog/display-a-beautiful-summary-statistics-in-r-using-skimr-package/)<p>
*   DataExplorer (https://www.programmingwithr.com/how-to-automate-eda-with-dataexplorer-in-r/)<p>


<strong>Posts:</strong><p>

*   My favourite R package for: summarising data (https://dabblingwithdata.amedcalf.com/2018/01/02/my-favourite-r-package-for-summarising-data/)<p>
*   How to Easily Create Descriptive Summary Statistics Tables in R Studio – By Group (https://thatdatatho.com/easily-create-descriptive-summary-statistic-tables-r-studio/)<p>
*   Twitter post by 'R-Ladies Jakarta' 'How to inspect your data set using R (https://twitter.com/RLadiesJakarta/status/1273236613410156546)<p>
*   Quick-R: Descriptive Statistics (https://www.statmethods.net/stats/descriptives.html)<p>
*   LittleMissData:  Explore Your Dataset in R (https://www.littlemissdata.com/blog/simple-eda)<p>


<strong>Visualizing Missing Data:</strong><p>

*   Gallery of Missing Data Visualizations (https://cran.r-project.org/web/packages/naniar/vignettes/naniar-visualisation.html)<p>


<strong>Graphs</strong><p>

*   The R Graph Gallery (https://r-graph-gallery.com/)<p>


